Type theories as quotient inductive-recursive types

### 2025-02-13 tex/main.tex
attempt to tighten Section 4

With the QIIR definition of a type theory, we show how $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ are eliminated to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}).
Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart (\cref{subsec:equivalence}).

@@

To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify types
\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
called the \emph{motivates} as the target types for the elimination principle:
\begin{alignat*}{5}
  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma &
  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \quad
               & \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
\end{alignat*}
Each constructor $c$ is interpreted by a corresponding functions $c^M$ indexed by $c$, e.g.,\ 
\begin{alignat*}{3}
  \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
  \blank,_{\Ctx}^M\blank & : \implicit{\dots}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) \\
  \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
  \emptysub_{\Sub}^M     & : \implicit{\dots}   && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
  \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
%  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
\end{alignat*}
and simultaneously defined functions are also interpreted by corresponding functions 
\begin{alignat*}{3}
  [\blank]_{\Ty}^M\blank & : \implicit{\dots} \; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) \\
  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) \\
  \blank\uparrow^M\blank & : \implicit{\dots} \; && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to
    \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)
\end{alignat*}
subject to the identities corresponding to equality constructors and function clauses, e.g., 
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\dots}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M \\
  [;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  [\idS]_{\Tm}^M & : \implicit{\dots}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} && t^M
\end{alignat*}
called \emph{methods}.
For brevity, we omit the rest of methods and refer to the Agda implementation for them.
\LT[noinline]{Add a link.}
We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$.

\begin{definition}
  An \emph{eliminator} for the QIIR definition of type theory is a tuple of motives and methods above.
  A \emph{recursor} is an eliminator whereas motives are constants over syntaxes.
\end{definition}

\begin{remark}
To state these methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory (e.g.\ $[\idS]_{\Tm}^M$) since identities (e.g.\ $[\idS]_{\Ty}^M$) are propositional just like equality constructors in its QII definition.
However, in the elimination principle for the QII definition, the situation is even worse: the above identity $[\idS]_{\Ty}^M$ already requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
\end{remark}

Given an eliminator, the actual computation for the elimination principle is given by
\begin{alignat*}{5}
  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) \\
  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
  \elim_{\Sub}\;\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
                            & \vdots
\end{alignat*}
interpreting each constructor using the corresponding method inductively.

Interestingly, the key difference from its QII counterpart is that $[\blank]_{\Ty}\blank$, $[\blank]_{\Tm}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
%type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
\begin{proposition}\label{prop:provable-function-clauses}
  Given the motives and the methods, we have
\begin{alignat*}{5}
  \elim_{\Ty}[] & : \elim_{\Ty}\,([\sigma]_{\Ty}\,A) && = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
  \nameless & : \elim_{\Tm}\,([\sigma]_{\Tm}\,t) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t) \\
  \nameless & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A) 
\end{alignat*}
\end{proposition}

These identities are needed to interpret constructors whereas strict identities are used in syntax.
Consider eliminating an extension $\sigma, t$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
\[
  \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,^M_{\Sub} \; \alert{t^M}.
\]
In the QII definition, $\elim_{\Tm}\,t$ is exactly $t^M$ and has the type
\begin{equation} \label{eq:elim-subext}
  \Tm^M\;\Gamma^M\;(\alert{\elim_{\Ty}\;([\sigma]_{\Ty}\,A)})\;t\judgeq
  \Tm^M\;\Gamma^M\;\left(\alert{[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)}\right)\;t
\end{equation}
where the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
In the QIIR definition, on the contrary, the interpreted type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$, which is not strictly equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
\[
  (\elim_{\Sub}\,\sigma) ,_{\Sub}^M \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
\]
transporting $\elim_{\Tm}\,t$ along the identity proof $\elim_{\Ty}[]$.

@@

\subsection{The standard model} \label{subsec:std-model}
The standard model, justifying the soundness of our definition, is given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
\begin{alignat*}{5}
  \Ctx^M & \defeq \lam{\nameless} \; && \Set & \qquad \Sub^M & \defeq \lam{\Gamma^M\,\Delta^M\,\nameless}\; && \Gamma^M \to \Delta^M \\
  \Ty^M  & \defeq \lam{\Gamma^M\, \nameless} && \Gamma^M \to \Set & 
  \Tm^M  & \defeq \lam{\Gamma^M\,A^M\,\nameless} && (\gamma : \Gamma^M) \to A^M\,\gamma
\end{alignat*}
with the additional methods to its QII counterpart:
\begin{align*}
  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M\,\gamma} (t^M\,(\sigma^M\,\gamma))  \\
                         & : (\sigma^M : \Gamma^M \to \Delta^M) (t^M : (\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
\end{align*}
where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.

In the QII counterpart, none of identities hold strictly: their indices always involve equality constructors.
E.g., the constructor $[\idS]_{\Ty}$ gives rise to
$[\idS]^M_{\Ty} : [ \idS^M ]^M_{\Ty}\, A^M =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]_{\Ty}}} A^M$
and the transport along $[\idS]_{\Ty}$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of~$A$.
Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs an extra step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport.
A workaround to avoid constant transports is to define the notion of recursors independent of eliminators and show that each recursor gives rise to an eliminator.

In contrast, in our QIIR definition, many identities just hold strictly.
Consider
\begin{alignat*}{5}
  [\idS]^M_{\Ty}                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
  \idS{\mathord{\uparrow}}^M & : \idS^M \uparrow^M A^M    && =^{\lam{B^M}\Sub^M\;(\Gamma^M, A^M)\;\Delta^M\;(\idS \uparrow A)}_{\alert{[\idS]^M}} \idS^M
\end{alignat*}
where $[\idS]^M_{\Ty}$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$.
By definition, it is a strict equality $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) \judgeq (\lam{\gamma}\gamma)$ by the $\eta$-law for $\Sigma$-types.

\subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
We have used additional rules to enable further reduction which are derivable in the QII definition.
Indeed, the QIIR definition of type theory is equivalent to its QII definition, as we can translate every constructor to its corresponding constructor and vice versa.
For comparison, we add a subscript to constructors and types in the QIIR and QII definitions by $\blank_\IR$ and $\blank_\I$, respectively, and write $(\blank)^{\succ}$ and $(\blank)^{\prec}$ for the translation from the QIIR definition to its QII counterpart and the converse, respectively.

The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QII counterpart as well as `smart constructors'.
Additional equality constructors are all derivable in the QII definition.

On the other hand, the converse $(\blank)^{\prec}$ has some choices.
As in the QII definition there is only one term substitution $[\blank]_{\I}\blank$ and lifting $(\blank)^+_{\I}$, we may choose to translate $[\sigma]_{\I} \,t$ to the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ to carry out some computation or the constructor form $[\sigma^{\prec}]_{\IR}\,t^{\prec}$ to stay verbatim. 
Either choice is possible, but we need to use \cref{prop:correctness-1,prop:correctness-3} to switch from one to the other whenever suitable.
For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $[ \sigma ]\,t = [\sigma]_{\Tm}\,t$ if the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ is chosen for the translation.
UIP is also needed,\footnote{%
We can further eliminate the use of propositional UIP by identifying function clauses with the reflexivity identity proof using additional rewrite rules.}
since the method $\mathsf{concat}^M_{\I}$ has a transport along $[;]_{\Ty}^M$ but $\mathsf{concat}_{\IR}$ does not.

>>>>

We show how to eliminate out of our QIIRT definition of type theory (\cref{subsec:eliminator}). In particular, we show how the elimination principle can be used to construct a standard model (\cref{subsec:std-model}) and compare our QIIRT definition with its QIIT counterpart (\cref{subsec:equivalence}).

@@

To eliminate $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$, we have to specify \emph{motive} type families
\begin{align*}
  \Ctx^M & : \Ctx \to \Set &
  \Sub^M & : \implicit{\Gamma, \Delta}\,\Ctx^M\;\Gamma \to \Ctx^M\;\Delta \to \Sub\;\Gamma\;\Delta \to \Set \\
  \Ty^M  & : \implicit{\Gamma}\,\Ctx^M\;\Gamma \to \Ty\;\Gamma \to \Set & 
  \Tm^M  & : \implicit{\Gamma, A}\; (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Tm\;\Gamma\;A \to \Set
\end{align*}
to act as the target types for the elimination principle:
\begin{alignat*}{5}
  \elim_{\Ctx} & : (\Gamma : \Ctx)   && \to \Ctx^M\;\Gamma &
  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\
  \elim_{\Ty}  & : (A : \Ty\;\Gamma) && \to \Ty^M\;(\elim_{\Ctx}\,\Gamma)\;A \quad
               & \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.
\end{alignat*}
Each constructor $c$ should be interpreted by a corresponding \emph{method} $c^M$ indexed by it, e.g.,
\begin{alignat*}{3}
%  \emptyctx_{\Ctx}^M     & :                    && \Ctx^M\;\emptyctx \\
  \blank,_{\Ctx}^M\blank & : \implicit{\Gamma, A}\; && (\Gamma^M : \Ctx^M\;\Gamma) \to \Ty^M\;\Gamma^M\;A \to \Ctx^M\;(\Gamma, A) %\\
%  \UU^M & : \implicit{\dots} && \Ty^M\;\Gamma^M\;\UU \\
%  \emptysub_{\Sub}^M     & : \implicit{\dots}   && \Sub^M\;\Gamma^M\;\emptyctx_{\Ctx}^M\;\cdot \\
%  \blank,_{\Sub}^M\blank & : \implicit{\dots}   && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Gamma^M\;([\sigma^M]^M_{\Ty}\,A^M)\;t \to \Sub^M\;\Gamma^M\;(\Delta^M ,^M_{\Ctx} A^M)\;(\sigma, t) \\
%  \idS^M          & : \implicit{\dots} && \Sub^M\;\Gamma^M\;\Gamma^M\;\idS \\
%  [\blank]^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]\,t)
\end{alignat*}
and simultaneously defined functions are also interpreted by corresponding functions, e.g.,
\begin{alignat*}{3}
  [\blank]_{\Ty}^M\blank & : \implicit{\Gamma, \Gamma^M,\Delta, \Delta^M, \sigma, A} \; && \Sub^M\;\Gamma^M\;\Delta^M\;\sigma \to \Ty^M\;\Delta^M\;A \to \Ty^M\;\Gamma^M\;([\sigma]_{\Ty}\,A) %\\
%  [\blank]_{\Tm}^M\blank & : \implicit{\dots} && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to \Tm^M\;\Delta^M\;A^M\;t \to \Tm^M\;\Gamma^M\;([\sigma^M]_{\Ty}^M\,A^M)([\sigma]_{\Tm}\,t) %\\
%  \blank\uparrow^M\blank & : \implicit{\dots} \; && (\sigma^M : \Sub^M\;\Gamma^M\;\Delta^M\;\sigma) \to (A^M : \Ty^M\Delta^M\;A) \to
%    \Sub^M\;(\Gamma^M, [\sigma^M]_{\Ty}^M\,A^M)\;(\Delta^M, A^M)
\end{alignat*}
subject to the identities corresponding to equality constructors and function clauses, e.g.,
\begin{alignat*}{5}
  [\idS]_{\Ty}^M & : \implicit{\Gamma, \Gamma^M, A, A^M}\; && [ \idS^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\;&& A^M %\\
  %[;]_{\Ty}^M    & : \implicit{\dots}   && [ \sigma^M ;^M \tau^M ]^M_{\Ty} \;A^M && =^{\Ty^M\,\Gamma^M}\; && [\sigma^M]^M_{\Ty}\,([\tau^M]^M_{\Ty}\,A^M) \\
  %[\idS]_{\Tm}^M & : \implicit{\Gamma^M, t, t^M}   && [\idS^M]^M_{\Tm}\;t^M && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} && \; t^M
\end{alignat*}
%
The elimination principle says that the functions $\elim_{\Ctx}$, $\elim_{\Sub}$, $\elim_{\Ty}$ and $\elim_{\Tm}$ as soon as motives and methods are given. When the motive families are constant, we call the principle a \emph{recursion principle} instead.
We refer to \cref{appendix:full-def} and our Agda development for the full definition of the elimination principle.
%We also have omitted the list of implicitly quantified variables: each variable $A$ implicitly quantified in a constructor $c$ has accordingly an $A$-indexed variable $A^M$ in $c^M$. 
%%
%
%\begin{definition}
%  An \emph{eliminator} for the QIIR definition of type theory is a tuple of motives and methods above.
%  A \emph{recursor} is an eliminator whereas motives are constants over syntaxes.
%\end{definition}
%

\begin{remark}
  To state the methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory --- for example, we have
  \[
    [\idS]_{\Tm}^M : \implicit{\Gamma^M, t, t^M} \;\;[\idS^M]^M_{\Tm}\;t^M  =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{[\idS]^M_{\Ty}}} \; t^M
  \]
  since identities occurring in a method %(e.g.\ $[\idS]_{\Ty}^M$)
  are propositional just like equality constructors in its QIIT definition.
  However, in the elimination principle for the QIIT definition, the situation is even worse, as we in general need yet another layer of transports for the methods to typecheck over the QIIT constructors.

  % : the above identity $[\idS]_{\Ty}^M$ already requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
%That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
\end{remark}

The elimination principle computes by interpreting each constructor using the corresponding method inductively, e.g.,
\begin{alignat*}{5}
%  \elim_{\Ctx}\;\emptyctx   & \reduce \emptyctx_{\Ctx}                                    && : \Ctx^M\;\emptyctx  \\
  \elim_{\Ctx}\;(\Gamma, A) & \reduce (\elim_{\Ctx}\,\Gamma),_{\Ctx}^{M} (\elim_{\Ty}\,A) && : \Ctx^M\;(\Gamma, A) %\\
%  \elim_{\Ty}\;\UU          & \reduce \UU^M && :\Ty^M\;\Gamma^M\;\UU \\
%  \elim_{\Sub}\;\emptysub   & \reduce \emptysub^M_{\Sub} &&: \Sub^M\;\Gamma^M\;\emptyctx^M_{\Ctx}\;\emptyctx \\
%                            & \vdots
\end{alignat*}
%
Interestingly, a key difference compared to its QIIT counterpart is that $[\blank]_{\Ty}\blank$, $[\blank]_{\Tm}$, and $\blank\uparrow\blank$ are functions rather than (formed by) constructors, meaning they cannot be eliminated inductively but instead become propositional identities.
%type substitution $[\blank]\blank$ was a constructor and $[\sigma]\,A$ was eliminated to $[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)$; similarly for term substitution.
\begin{proposition}\label{prop:provable-function-clauses}
  Given the motives and the methods, we can prove
\begin{alignat*}{5}
  \elim_{\Ty}[] & : \elim_{\Ty}\,([\sigma]_{\Ty}\,A) && = && [\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A) \\
  \elim_{\Tm}[] & : \elim_{\Tm}\,([\sigma]_{\Tm}\,t) && =^{\lam{A^M}\Tm^M\;\Gamma^M\;A^M\;t}_{\alert{\elim_{\Ty}[]}} && [\elim_{\Sub}\,\sigma]^M_{\Tm}\,(\elim_{\Tm}\,t) \\
  \elim_{\Sub}[] & : \elim_{\Sub}(\sigma\uparrow A) && =^{\lam{B^M}\Sub^M\;(\Gamma^M,^M B^M)\;(\Delta^M, A^M)\;(\sigma \uparrow A)}_{\alert{\elim_{\Ty}[]}} && (\elim_{\Sub} \sigma) \uparrow^M (\elim_{\Ty}\,A) 
\end{alignat*}
\end{proposition}

These identities are needed for the computation rules where strict identities are used in syntax.
For example, for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, we have 
\[
  \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,_{\Sub}^M \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
\]
where the transport is needed since $\elim_{\Tm}\;t : \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\alert{\elim_{\Ty}\;([\sigma]_{\Ty}\,A)})\;t$ but $\blank ,_{\Sub}^M \blank$ expects its second argument to have type $\Tm^M\;(\elim_{\Ctx}\,\Gamma)\;\left(\alert{[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)}\right)\;t$.

% Consider eliminating a substitution extension $\sigma, t$ for $\sigma : \Sub\;\Gamma\;\Delta$ and $t : \Tm\;\Gamma\;([\sigma]_{\Ty}\,A)$, i.e.\ 
% \[
%   \elim_{\Sub}(\sigma, t) = (\elim_{\Sub}\,\sigma) ,^M_{\Sub} \; \alert{t^M}.
% \]
% In the QIIT definition, $\elim_{\Tm}\,t$ is exactly $t^M$ and has the type
% \begin{equation} \label{eq:elim-subext}
%   \Tm^M\;\Gamma^M\;(\alert{\elim_{\Ty}\;([\sigma]_{\Ty}\,A)})\;t\judgeq
%   \Tm^M\;\Gamma^M\;\left(\alert{[\elim_{\Sub}\,\sigma]^M_{\Ty}\,(\elim_{\Ty}\,A)}\right)\;t
% \end{equation}
% where the RHS of \eqref{eq:elim-subext} is indeed the type of the second argument for $\blank,^M_{\Sub}\blank$.
% In the QIIRT definition, on the contrary, the interpreted type of $\elim_{\Tm}\;t$ is $\elim_{\Ty}\;([\sigma]_{\Ty}\,A)$, which is not strictly equal to the RHS of \eqref{eq:elim-subext}, so $\elim_{\Sub}(\sigma, t)$ has to be
% \[
%   (\elim_{\Sub}\,\sigma) ,_{\Sub}^M \alert{\transfib{\Tm^M\;(\elim_{\Ctx}\,\Gamma)}{\elim_{\Ty}[]}{\color{black}(\elim_{\Tm}\,t)}}
% \]
% transporting $\elim_{\Tm}\,t$ along the identity proof $\elim_{\Ty}[]$.

@@

\subsection{The standard model} \label{subsec:std-model}
The standard model, justifying the soundness of our definition, is given by the same motives as those for its QIIT counterpart~\cite[Section~4]{Altenkirch2016a}:
\begin{alignat*}{5}
  \Ctx^M & \defeq \lam{\nameless} \; && \Set & \qquad \Sub^M & \defeq \lam{\Gamma^M\,\Delta^M\,\nameless}\; && \Gamma^M \to \Delta^M \\
  \Ty^M  & \defeq \lam{\Gamma^M\, \nameless} && \Gamma^M \to \Set & 
  \Tm^M  & \defeq \lam{\Gamma^M\,A^M\,\nameless} && (\gamma : \Gamma^M) \to A^M\,\gamma
\end{alignat*}
%with the additional methods to its QII counterpart:
%\begin{align*}
%  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
%                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
%  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M\,\gamma} (t^M\,(\sigma^M\,\gamma))  \\
%                         & : (\sigma^M : \Gamma^M \to \Delta^M) (t^M : (\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
%\end{align*}
%where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.
%
We once again refer to our Agda formalisation for details.
We remark that because our lack of transports in the syntax, many of the identities in the standard model simply hold strictly --- more so than in the QIIT standard model, where an extra step~\cite[Lemma~2.3.5]{UFP2013} is often needed for reasoning about constant transports.

%In the QIIT counterpart standard model, none of identities hold strictly, since their indices always involve equality constructors.
%E.g., the constructor $[\idS]_{\Ty}$ gives rise to
%$[\idS]^M_{\Ty} : [ \idS^M ]^M_{\Ty}\, A^M =^{\Ty^M\;\Gamma^M}_{\alert{[\idS]_{\Ty}}} A^M$
%and the transport along $[\idS]_{\Ty}$ is needed, because $[ \idS^M ]^M_{\Ty}\, A^M$ is indexed by $[\idS]\,A$ instead of~$A$.
%Accordingly, this identity becomes $\lam{\gamma} A^M\,\gamma =_{\alert{[\idS]}}^{\lam{\nameless}\Gamma^M \to \Set} A^M$ in the standard model and needs an extra step~\cite[Lemma~2.3.5]{UFP2013} for reasoning about the constant transport.
%A workaround to avoid constant transports is to define the notion of recursors independent of eliminators and show that each recursor gives rise to an eliminator. 
%
%In contrast, in our QIIR definition, many identities just hold strictly.
%Consider
%\begin{alignat*}{5}
%  [\idS]^M_{\Ty}                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
%  \idS{\mathord{\uparrow}}^M & : \idS^M \uparrow^M A^M    && =^{\lam{B^M}\Sub^M\;(\Gamma^M, A^M)\;\Delta^M\;(\idS \uparrow A)}_{\alert{[\idS]^M}} \idS^M
%\end{alignat*}
%where $[\idS]^M_{\Ty}$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
%Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$.
%By definition, it is a strict equality $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) \judgeq (\lam{\gamma}\gamma)$ by the $\eta$-law for $\Sigma$-types.

\subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
We have used additional rules to enable further reduction which are derivable in the QIIT definition.
Indeed, the QIIRT definition of type theory is equivalent to its QIIT definition, as we can translate every constructor to its corresponding constructor and vice versa.
For comparison, we add a subscript to constructors and types in the QIIRT and QIIT definitions by $\blank_\IR$ and $\blank_\I$, respectively, and write $(\blank)^{\succ}$ and $(\blank)^{\prec}$ for the translation from the QIIRT definition to its QIIT counterpart and the converse, respectively.

The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QIIT counterpart as well as `smart constructors'.
Additional equality constructors are all derivable in the QIIT definition.
